Realizing temporal graphs from fastest travel times

### 2023-03-21 New Paper/Realizing TG - extended.tex
New paper

\usepackage{caption} %for subfigure - join multiple figures and add captions
%\usepackage{subcaption}
\usepackage{changepage} %inside figure we can put \begin{adjustwidth}{-1cm}{-1cm} \end{adjustwidth} and we can ignore the margins
\usepackage{mathtools} % write text [under]{over} arrow $\xrightarrow[\text{world}]{\text{hello}}$

@@

\bibliographystyle{plainurl}% the mandatory bibstyle
\usepackage{cite} %order citations by increasing number
\title{Realizing temporal graphs from fastest travel times} %TODO Please add

@@

\begin{document}
\maketitle

\begin{abstract}
In this paper we initiate the study of the \emph{temporal graph realization} problem with respect to the fastest path durations among its vertices, 
while we focus on periodic temporal graphs. 
Given an $n \times n$ matrix $D$ and a $\Delta \in \mathbb{N}$, the goal is to construct a $\Delta$-periodic temporal graph with $n$ vertices 
such that the duration of a \emph{fastest path} from $v_i$ to $v_j$ is equal to $D_{i,j}$, or to decide that such a temporal graph does not exist. 
The variations of the problem on static graphs has been well studied and understood since the 1960's, and this area of research remains active until nowadays.

As it turns out, the periodic temporal graph realization problem has a very different computational complexity behavior than its static (i.e.~non-temporal) counterpart. 
First we show that the problem is NP-hard in general, but polynomial-time solvable if the so-called underlying graph is a tree or a cycle.
Building upon those results, we investigate its parameterized computational complexity with respect to structural parameters of the underlying static graph which measure the ``tree-likeness''. For those parameters, we essentially give a tight classification between parameters that allow for tractability (in the FPT sense) and parameters that presumably do not.
We show that our problem is W[1]-hard when parameterized by the \emph{feedback vertex number} of the underlying graph, while we show that it is in FPT when parameterized by the \emph{feedback edge number} of the underlying graph.

\vspace{0,2cm}
\noindent \textbf{Due to lack of space, the full paper with all proofs is attached in a 
	clearly marked Appendix to be read at the discretion of the Program Committee.}
\end{abstract}

\section{Introduction}\label{intro-sec}

The \emph{graph realization} problem with respect to a graph property $\mathcal{P}$ is to find a graph that satisfies property $\mathcal{P}$, or to decide that no such graph exists. 
The motivation for graph realization problems stems both from ``verification'' applications, 
where we attempt to reconstruct an existing network of unknown structure based on the outcomes of experimental measurements, 
and from network design applications in engineering, where the goal is to design network topologies having a desired property~\cite{augustine2022distributed,grotschel1995design}.
Analyzing the computational complexity of the graph realization problems for various natural and fundamental graph properties $\mathcal{P}$ requires a deep understanding of these properties.
Among the most well studied such parameters for graph realization 
are constraints on the distances between vertices~\cite{barNoy2022GraphRealization,barNoy2021composed,hakimi1965distance,chung2001distance,bixby1988almost,culberson1989fast}, 
on the vertex degrees~\cite{GolovachM17,gomory1961multi,hakimi1962realizability,Bar-NoyCPR20,erdos1960graphs}, 
%%%%barNoy2021selected,cai2000solution
on the eccentricities~\cite{barNoy2020efficiently,hell2009linear,behzad1976eccentric,lesniak1975eccentric}, and on connectivity~\cite{fulkerson1960zero,frank1992augmenting,chen1966realization,frank1994connectivity,frank1970connectivity,gomory1961multi}, among others.

In the simplest version of a graph realization problem with respect to vertex distances, 
we are given a symmetric $n \times n$ matrix $D$ and we are looking for an $n$-vertex undirected and unweighted graph $G$ such that $D_{i,j}$ equals the distance between vertices $v_i$ and $v_j$ in~$G$. This problem can be trivially solved in polynomial time in two steps~\cite{hakimi1965distance}: First, we build the graph $G=(V,E)$ such that $v_i v_j \in E$ if and only if $D_{i,j}=1$. Second, from this graph $G$ we compute the matrix $D_G$ which captures the shortest distances for all pairs of vertices. If $D_G = D$ then $G$ is the desired graph, otherwise there is no graph having $D$ as its distance matrix. 
Non-trivial variations this problem have been extensively studied, such as for weighted graphs~\cite{hakimi1965distance,Patrinos-Hakimi-72}, as well as for cases where the realizing graph has to belong to a specific graph family~\cite{hakimi1965distance, barNoy2021composed}. Other variations of the problem include the cases where every entry of the input matrix $D$ may contain a range of consecutive permissible values~\cite{barNoy2021composed,Rubei16,Tamura93}, or even an arbitrary set of acceptable values~\cite{barNoy2022GraphRealization} for the distance between the corresponding two vertices.

In this paper we make the first attempt to understand the complexity of the graph realization problem with respect to vertex distances in the context of \emph{temporal graphs}, i.e.~of graphs whose \emph{topology changes over time}.

\begin{definition}[temporal graph~\cite{KKK00}]
\label{temp-graph-def} A \emph{temporal graph} is a pair $(G,\lambda)$,
where $G=(V,E)$ is an underlying (static) graph and $\lambda :E\rightarrow 2^\mathbb{N}$ is a \emph{time-labeling} function which assigns to every edge of $G$ a set of discrete time-labels.
\end{definition}

Here, whenever $t \in \lambda(e)$, we say that the edge $e$ is \emph{active} or \emph{available} at time $t$. In the context of temporal graphs, where the notion of vertex adjacency is time-dependent, the notions of path and distance also need to be redefined. The most natural temporal analogue of a path is that of a \emph{temporal} (or \emph{time-dependent}) path, which is motivated by the fact that, due to
causality, entities and information in temporal graphs can ``flow'' only along sequences of
edges whose time-labels are strictly increasing.

\begin{definition}[fastest temporal path] \label{def:temporalPath+Duration}
Let $(G,\lambda)$ be a temporal graph. A \emph{temporal path} 
in $(G,\lambda)$ is a sequence $(e_1,t_1),(e_2,t_2),\ldots,(e_k,t_k)$, 
where $P=(e_1,\ldots,e_k)$ is a path in the underlying static graph $G$, 
%(we usually say that $P$ is the underlying path of $(P,\lambda)$), 
$t_i\in \lambda(e_i)$ for every $i=1,\ldots,k$, and $t_1<t_2<\ldots<t_k$. 
The \emph{duration} of this temporal path %$(P,\lambda)$ from $u$ to $v$ 
is $t_k - t_1 + 1$.
A \emph{fastest} temporal path from a vertex $u$ to a vertex $v$ in $(G,\lambda)$ is a temporal path from $u$ to $v$ with the smallest duration.
The duration of the \emph{fastest} temporal path from $u$ to $v$ is denoted by $d(u,v)$.
\end{definition}

%Other temporal analogues of a shortest path are temporal paths with the smallest total number of edges (\emph{shortest} temporal path), or the smallest earliest arrival time (\emph{foremost} temporal path), respectively.

In an arbitrary temporal graph $(G,\lambda)$ there can be pairs of vertices $u,v$ where there is no temporal path at all from $u$ to $v$, even if the underlying graph $G$ is undirected and connected. 
In order to focus on the smallest duration of temporal graphs, rather than on temporal reachability between vertices, we consider in this paper only \emph{periodic} temporal graphs, i.e.~on temporal graphs in which the availability times of every edge are periodic (all edges having the same period $\Delta$)~\cite{casteigts2012time}. 
Apart from the theoretical interest of periodic temporal graphs, which have not been well studied and understood in the literature, they capture naturally many application areas too, such as transportation or satellite networks, where entities move according to repeating schedules or time tables. For example, a fastest temporal path from vertex $u$ to vertex $v$ in the temporal graph defined by a transportation (resp.~satellite) network represents the shortest travel time (resp.~the shortest communication delay) from $u$ to $v$. 
As it turns out, the periodic temporal graph realization problem with respect to a given $n \times n$ matrix $D$ of the fastest duration times has a very different computational complexity behavior than the classic graph realization problem with respect to shortest path distances in static graphs.

Formally, let $G=(V,E)$ and $\Delta\in \mathbb{N}$, and let $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ be an edge-labeling function that assigns to every edge of $G$ exactly one of the labels from $\{1,\ldots,\Delta\}$. 
Then we denote by $(G,\lambda,\Delta)$ the \emph{$\Delta$-periodic temporal graph} $(G,L)$, where for every edge $e\in E$ we have $L(e)=\{i\Delta + x : i\geq 0, x\in \lambda(e)\}$. 
In this case we call $\lambda$ a \emph{$\Delta$-periodic labeling} of $G$. 
When it is clear from the context, we drop $\Delta$ form the notation and say that $(G,\lambda)$ is a ($\Delta$-periodic) temporal graph.
Given a duration matrix $D$, it is easy to observe that similar to the static case, if $D_{i,j}=1$, then $v_i$ and $v_j$ must be connected by an edge. We call the graph defined by those edges as the \emph{underlying graph} of $D$.

\subparagraph{Our contribution.}
%In this work, 
We initiate the study of naturally motivated graph realization problems in the temporal setting. To this end, we introduce and investigate the following problem, capturing the setting described above:

\problemdef{\deltaExactLong\ (\deltaExact)}
{An $n \times n$ matrix $D$, a positive integer $\Delta$.}
{Does there exist a graph $G=(V,E)$ with vertices $\{v_1,\ldots,v_{n}\}$ 
and a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ such that, 
for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is $D_{i,j}$?}

We start with showing NP-hardness of the problem (\cref{thm:NPhardness}), even if $\Delta$ is a small constant. To establish a baseline for tractability, we show that \deltaExact\ is polynomial-time solvable if the underlying graph is a tree (\cref{thm:deltaExact-PolyTimeTrees}).

Building upon these initial results, we explore the possibilities to generalize our polynomial-time algorithm using the \emph{distance-from-triviality} parameterization paradigm~\cite{FJR13,GHN04}. That is, we investigate the parameterized computational complexity of \deltaExact\ with respect to structural parameters of the underlying graph that measure its ``tree-likeness''.

We obtain the following results. We show that \deltaExact\ is W[1]-hard when parameterized by the feedback vertex number of the underlying graph (\cref{thm:W1wrtFVS}). 
To this end, we first give a reduction from \textsc{Multicolored Clique} parameterized by the number of colors~\cite{fellows2009multipleinterval} to a variant of \deltaExact\ where the period $\Delta$ is infinite, that is, when the labeling is non-periodic. We use a special gadget (the ``infinity'' gadget) which allows us to transfer the result to a finite period $\Delta$. The latter construction is independent from the particular reduction we use, and can hence be treated as a reduction from the non-periodic to the periodic setting.

We complement this result by showing that the problem is fixed-parameter tractable (FPT) with respect to the feedback edge number of the underlying graph (\cref{thm:FPTwrtFES}). 
Our FPT algorithm preprocesses the input graph in a suitable way that allows us to efficiently guess desired structures. Then we create integer linear programs (ILPs) (with bounded number of variables) and add constraints to those instances by suitably partitioning the fastest paths among pairs of vertices belonging to a small number of interesting sets. The notion of segment is crucial for adding the necessary and sufficient number of linear constraints. We use Lenstra's Theorem~\cite{Lenstra1983Integer} to solve the ILPs efficiently.

Note that this provides a fairly complete picture of the parameterized computational complexity of \deltaExact\ with respect to structural parameters of the underlying graph that measure ``tree-likeness''. To obtain our results, we prove several properties of fastest temporal paths, which may be of independent interest.

Due to space constraints, proofs of results marked with $\star$ are (partially) deferred to the Appendix.

\subparagraph{Related work.} Graph realization problems on static graphs have been studied since the 1960s. We provide an overview of the literature in the introduction.

To the best of our knowledge, we are the first to consider graph realization problems in the temporal setting. However, many other connectivity-related problems have been studied in the temporal setting~\cite{enright2021deleting,MolterRZ21,klobas2023interference,deligkas2022optimizing,erlebach2021temporal,Flu+19a,Zsc+19,CasteigtsCS22}, most of which are much more complex and computationally harder than their non-temporal counterparts, and some of which do not even have a non-temporal counterpart.

There are two problem settings that share similarities with our setting, which we discuss in more detail in the following.

Several problems have been studied where the goal is to assign labels to (sets of) edges of a given static graph in order to achieve certain connectivity-related properties~\cite{KlobasMMS22,MertziosMS19,akrida2017complexity,enright2021assigning}. The main difference to our problem setting is that in the mentioned works, the input is a graph and the sought labeling is not periodic. Furthermore, the investigated properties are temporal connectivity among all vertices~\cite{KlobasMMS22,MertziosMS19,akrida2017complexity}, temporal connectivity among a subset of vertices~\cite{KlobasMMS22}, or reducing reachability among the vertices~\cite{enright2021assigning}. In all cases the duration of the temporal connections is not of interest.

The only work we are aware of that studies problems on periodic temporal graphs is work on the temporal cops and robber game~\cite{Morawietz021,MorawietzRW20,ErlebachS20}. The question here is (in its basic case) whether a certain number of cops are able to catch a robber, if everybody can move along one edge at every time step and the robber is considered caught if it occupies the same vertex as a cop at the same time.

\subparagraph{Preliminaries and notation.}
We already introduced the most central notion and concepts. There are some additional definitions we need, to present our proofs and results which we give in the following.

An interval in $\mathbb N$ from $a$ to $b$ is denoted by $[a,b] = \{ i\in \mathbb N  :  a \leq i \leq b\}$; similarly, $[a] = [1,a]$.
An undirected graph~$G=(V,E)$ consists of a set~$V$ of vertices 
and a set~$E \subseteq V \times V$ of edges.
For a graph~$G$, we also denote by~$V(G)$ and~$E(G)$ the vertex and edge set of~$G$, respectively.
We denote an edge $e \in E$ between vertices $u,v \in V$ as a set $e=\{u,v\}$.
For the sake of simplicity of the representation, an edge $e$ is sometimes also denoted by $uv$. 
A path~$P$ in $G$ is a subgraph of $G$ with vertex set~$V(P)=\{v_1,\dots,v_k\}$ and edge set~$E(P)=\{\{v_i,v_{i+1}\} :  1\leq i<k\}$
(we often represent path~$P$ by the tuple~$(v_1,v_2,\dots,v_k)$).

Let $v_1,v_2,\ldots,v_n$ be the $n$ vertices of the graph $G$. 
For simplicity of the presentation~--and with a slight abuse of notation--~we refer during the paper to the entry $D_{i,j}$ of the matrix $D$ as 
$D_{a,b}$, where $a=v_i$ and $b=v_j$. That is, we put as indices of the matrix $D$ the corresponding vertices of $G$ whenever it is clear from the context.
Many times we also refer to a path $P=(u=v_1, v_2, \dots, v_p=v)$ from $u$ to $v$ in $G$,
as a temporal path in $(G,\lambda)$,
where we actually mean that $(P,\lambda,\Delta)$ is a temporal path with $P$ as its underlying (static) path.

We remark that a fastest path between two vertices in a temporal graph can be computed in polynomial time~\cite{xuan_computing_2003,Wu2016Efficient}.
%Wu et~al.~\cite{Wu2016Efficient} provide an algorithm that computes the duration of the fastest temporal path from a source vertex $v$, in a temporal graph $(G,\lambda)$, to all other vertices,
%that runs in polynomial-time.
Hence, given a $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$, we can compute in polynomial-time  the matrix $D$
which consists of durations of fastest temporal paths among all pairs of vertices in $(G,\lambda,\Delta)$.

\begin{comment}
Let $(G,\lambda)$ be a temporal graph with  $n = |V(G)|$ and  $m=|E(G)|$.
Let $v \in V(G)$ be an arbitrary vertex of $G$, denote with $deg_{max}$ the maximum degree of all vertices in $V(G)$ and let $S_v$ be the set of time edges in $(G,\lambda)$, incident to $v$, \ie $S_v = \{(uv, \lambda(uv)) | uv \in E(G)\}$.
In their work Wu et~al.~\cite{Wu2016Efficient} provide an algorithm that computes the duration of the fastest temporal path from a source vertex $v$ to all other vertices in $G$ in
$O(n + m \log c)$ time, where $c = \min \{|S_v|, deg_{max}\}$.
From this result, we can deduce the following. 
\begin{lemma} \label{lemma:calculatingD}
Let $(G, \lambda, \Delta)$ be a $\Delta$-periodic temporal graph, where $\lambda$ assigns exactly $k$ labels to every edge of $G$. 
Denote with $deg_{max}$ the maximum degree of vertices in $G$, let $deg(v_i)$ be the degree of vertex $v_i \in V(G)$,
and let $c_{v_i} = \min \{ k \cdot deg(v_i), deg_{max}\}$.\\
To calculate the durations of all fastest temporal paths from a fixed vertex $v \in V(G)$ to all other vertices, it takes $O(n + m \log c_v)$ time. \\
To calculate the durations of the fastest temporal paths among all pairs of vertices we then need $O(n^2 + m \sum_{v \in V(G)} \log c_v)$ time.
\end{lemma}

Observe that $O(n + m \log c_v) = O(n + m \log deg_{max}) = O(n + m \log n) = O(n^2 \log n)$ and similarly
$O(n^2 + m \sum_{v \in V(G)} \log c_v) = O(n^2 + mn \log deg_{max}) = O(n + mn \log n) = O(n^3 \log n)$.
Since an instance $I$ of \kDeltaExact\ problem is an $n \times n$ matrix $D$, the running time of calculating one row of $D$ (\ie the fastest temporal paths from a single vertex to all others) takes $O(|I|\log \sqrt{|I|})$-time and calculating the whole matrix $D$ (\ie the fastest temporal paths among all pairs of vertices) takes $O(|I|^{3/2} \log \sqrt{|I|})$ time.
Therefore, from now on we say that calculating (a row of) $D$ takes polynomial time.
\end{comment}
%\section{Results}

\section{Hardness results for \deltaExact}  
In this section we present our main computational hardness results.  
We first show that \deltaExact\ is NP-hard even for constant $\Delta$. 
%Then we investigate the parameterized computational hardness of \deltaExact\ with respect to structural parameters of the underlying graph. We show that \deltaExact\ is W[1]-hard when parameterized by the feedback vertex number of the underlying graph.

%\subparagraph{NP-hardness of \deltaExact.}\label{sec:nphardness}
%We prove that in general it is NP-hard to determine a $\Delta$-periodic temporal graph $(G,\lambda)$ respecting a duration matrix $D$,
%even if $\Delta$ is a small constant.

\begin{theorem}[$\star$]\label{thm:NPhardness}
	\deltaExact\ is \NP-hard for all $\Delta \geq 3$.
\end{theorem}

\begin{comment}

\begin{proof}
	We present a polynomial-time reduction from the NP-hard problem NAE 3-SAT~\cite{Schaefer1978complexity}. Here we are given a formula $\phi$ that is a conjunction of so-called NAE (not-all-equal) clauses, where each clause contains exactly 3 literals (with three distinct variables).
    A NAE clause evaluates to \textsc{true} if and only if not all of its literals are equal, that is, at least one literal evaluates to \textsc{true} and at least one literal evaluates to \textsc{false}.
    We are asked whether $\phi$ admits a satisfying assignment.

    Given an instance $\phi$ of NAE 3-SAT, we construct an instance $(D,\Delta)$ of \deltaExact\ as follows.

 We start by describing the vertex set of the underlying graph $G$ of $D$.
\begin{itemize}
\item For each variable $x_i$ in $\phi$, we create three variable vertices $x_i, x_i^T, x_i^F$.
\item For each clause $c$ in $\phi$, we create one clause vertex $c$.
\item We add one additional super vertex $v$.
\end{itemize}
Next, we describe the edge set of $G$.
\begin{itemize}
\item For each variable $x_i$ in $\phi$ we add the following five edges: 
$x_i x_i^T, x_i x_i^F, x_i^T x_i^F, x_i^T v$ and $ x_i^Fv$.
\item For each pair of variables $x_i,x_j$ in $\phi$ with $i \neq j$ we add the following four edges: 
$x_i^T x_j^T, x_i^Tx_j^F, x_i^Fx_j^T$ and $x_i^Fx_j^F$.
\item For each clause $c$ in $\phi$ we add one edge for each literal. Let $x_i$ appear in $c$. If $x_i$ appears non-negated in $c$ we add edge $c x_i^T$. 
If $x_i$ appears negated in $c$ we add edge $c x_i^F$.
\end{itemize}
This finishes the construction of $G$.
For an illustration see~\cref{fig:NP-example}.

%
\begin{figure}[htb]
	\noindent
	\makebox[\textwidth]{
		\centering
\begin{tikzpicture}%[xscale=2]

	%%Variable X1
	\node (x1t) at (1,0) {};
	\node (x1f) at (3,0) {};

	%%Variable X2
	\node (x2t) at (5,0) {};
	\node (x2f) at (7,0) {};

	%%Variable x3
	\node (x3t) at (9,0) {};
	\node (x3f) at (11,0) {};

	%%Variable x4
	\node (x4t) at (13,0) {};
	\node (x4f) at (15,0) {};

	%% edges among xt, xf vertices
	%x1 to all
	\draw [dotted,gray] (x1t) [out=30,in=150] to (x2t);
	\draw [dotted,gray] (x1t) [out=30,in=150] to (x3t);
	\draw [dotted,gray] (x1t) [out=30,in=150] to (x4t);
	\draw [dotted,gray] (x1t) [out=30,in=150] to (x2f);
	\draw [dotted,gray] (x1t) [out=30,in=150] to (x3f);
	\draw [dotted,gray] (x1t) [out=30,in=150] to (x4f);

	\draw [dotted,gray] (x1f) [out=30,in=150] to (x2t);
	\draw [dotted,gray] (x1f) [out=30,in=150] to (x3t);
	\draw [dotted,gray] (x1f) [out=30,in=150] to (x4t);
	\draw [dotted,gray] (x1f) [out=30,in=150] to (x2f);
	\draw [dotted,gray] (x1f) [out=30,in=150] to (x3f);
	\draw [dotted,gray] (x1f) [out=30,in=150] to (x4f);
	%x2 to all
	\draw [dotted,gray] (x2t) [out=30,in=150] to (x3t);
	\draw [dotted,gray] (x2t) [out=30,in=150] to (x4t);
	\draw [dotted,gray] (x2t) [out=30,in=150] to (x3f);
	\draw [dotted,gray] (x2t) [out=30,in=150] to (x4f);

	\draw [dotted,gray] (x2f) [out=30,in=150] to (x3t);
	\draw [dotted,gray] (x2f) [out=30,in=150] to (x4t);
	\draw [dotted,gray] (x2f) [out=30,in=150] to (x3f);
	\draw [dotted,gray] (x2f) [out=30,in=150] to (x4f);
	%x3 to all
	\draw [dotted,gray] (x3t) [out=30,in=150] to (x4t);
	\draw [dotted,gray] (x3t) [out=30,in=150] to (x4f);

	\draw [dotted,gray] (x3f) [out=30,in=150] to (x4t);
	\draw [dotted,gray] (x3f) [out=30,in=150] to (x4f);

	%%Variable X1
	\node[vert,label=left:$x_1^T$] (x1t) at (1,0) {};
	\node[vert,label={[fill=white]right:$x_1^F$}] (x1f) at (3,0) {};
	\node[vert,label=above:$x_1$] (x1) at (2,1) {};
	\draw (x1t) -- (x1f) -- (x1) -- (x1t);

	%%Variable X2
	\node[vert,label={[fill=white]left:$x_2^T$}] (x2t) at (5,0) {};
	\node[vert,label={[fill=white]right:$x_2^F$}] (x2f) at (7,0) {};
	\node[vert,label={[fill=white]above:$x_2$}] (x2) at (6,1) {};
	\draw (x2t) -- (x2f) -- (x2) -- (x2t);

	%%Variable x3
	\node[vert,label={[fill=white]left:$x_3^T$}] (x3t) at (9,0) {};
	\node[vert,label={[fill=white]right:$x_3^F$}] (x3f) at (11,0) {};
	\node[vert,label={[fill=white]above:$x_3$}] (x3) at (10,1) {};
	\draw (x3t) -- (x3f) -- (x3) -- (x3t);

	%%Variable x4
	\node[vert,label={[fill=white]left:$x_4^T$}] (x4t) at (13,0) {};
	\node[vert,label=right:$x_4^F$] (x4f) at (15,0) {};
	\node[vert,label=above:$x_4$] (x4) at (14,1) {};
	\draw (x4t) -- (x4f) -- (x4) -- (x4t);

	%%vertices for clauses c1, c2 + vertex v
	\node[vert,label=above:$v$] (v) at (8,3) {};
	\node[vert,label=below:$c_1$] (c1) at (4.5,-3) {};
	\node[vert,label=below:$c_2$] (c2) at (11.5,-3) {};

	%%% edges between x1 gadget and v,c1,c2
	\draw (x1t) to node[timelabel] {$1$} (c1);
	\draw (x1t) to node[timelabel] {$1$} (c2);
	\draw (x1t) to [out=90,in=190] (v);
	\draw (x1f) to [out=90,in=200] node[timelabel] {$3$} (v);

	%%% edges between x2 gadget and v,c1,c2
	\draw (x2f) to node[timelabel] {$1$} (c1);
	\draw (x2t) to node[timelabel] {$1$} (c2);
	\draw (x2t) to [out=80,in=210]  (v);
	\draw (x2f) to node[timelabel] {$3$} (v);

	%%% edges between x3 gadget and v,c1
	\draw (x3t) to node[timelabel] {$1$} (c1);
	\draw (x3t) to node[timelabel] {$3$} (v);
	\draw (x3f) to [out=100,in=330] (v);

	%%% edges between x4 gadget and v,c2
	\draw (x4t) to node[timelabel] {$1$} (c2);
	\draw (x4t) to [out=90,in=340] node[timelabel] {$3$} (v);
	\draw (x4f) to [out=90,in=360] (v);

\end{tikzpicture}
}
\caption{Illustration of the temporal graph $(G,\lambda)$ from the NP-hardness reduction, 
	where the NAE 3-SAT formula $\phi$ is of form $\phi = \text{NAE}(x_1, \overline{x}_2, x_3) \wedge \text{NAE}(\overline{x}_1, x_2, x_4)$.
To improve the readability, we draw edges between vertices $x_i^T$ and $x_j^F$ (where $i \neq j$) with gray dotted lines.
Presented is the labeling of $G$ corresponding to the assignment $x_1=x_2=\textsc{true}$ and $x_3,x_4=\textsc{false}$,
where all unlabeled edges get the label $2$.
}\label{fig:NP-example}
\end{figure}
%
We set $\Delta$ to some constant larger than two, that is, $\Delta\ge 3$. Next, we specify the durations in the matrix $D$ between all vertex pairs.
For the sake of simplicity we write $D_{u,v}$ as $d(u,v)$,
where $u,v$ are two vertices of $G$. 
We start by setting the value of $d(u,v) = 1$ where $u$ and $v$ are two adjacent vertices in $G$.
\begin{itemize}
    %\item For each variable $x$ in $\phi$ we specify the following distances between the non-adjacent variable vertices:
    %$d(x_1,x_2)=2$. 
    %\item For pair of variable $x,y$ in $\phi$ with $x\neq y$ we specify the following 
    %$d(x_1,y^T)=d(x_1,y^F)=3$, $d(x_2,y^T)=d(x_2,y^F)=2$.
    \item For each variable $x_i$ in $\phi$ and the super vertex $v$
    we specify the following durations:
    $d(x_i,v)=2$ and $d(v,x_i)= \Delta $. %, $d(x_2,v)=3$.
    \item For each clause $c$ in $\phi$ and the super vertex $v$
    we specify the following durations:
    $d(c,v)=2$ and $d(v,c)= \Delta - 1$.
    \item Let $x_i$ be a variable that appears in clause $c$, then  we specify the following durations:
    $d(c,x_i)=2$ and $d(x_i,c)=\Delta$.
    If $x_i$ appears non-negated in $c$ we specify the following durations:
    $d(c,x_i^F)=2$ and $d(x_i^F,c)=\Delta$.
    If $x_i$ appears negated in $c$ we specify the following duratios:
    $d(c,x_i^T)=2$ and $d(x_i^T,c)=\Delta$.
    \item Let $x_i$ be a variable that does \emph{not} appear in clause $c$, then we specify the following duratios:
    $d(x_i,c)=\Delta$, $d(c,x_i)=2\Delta + 2$
    and
    $d(c,x_i^T)=d(c,x_i^F)=2$, $d(x_i^T,c)=d(x_i^F,c)=\Delta$. 
    \item For each pair of variables $x_i \neq x_j$ in $\phi$ we specify the following duratios:
    $d(x_i,x_j)=2\Delta +1$ and
    $d(x_i,x_j^T)=d(x_i,x_j^F)=\Delta + 1$.
    \item For each pair of clauses $c_i \neq c_j$ in $\phi$ we specify the following duratios:
    $d(c_i,c_j)= \Delta + 1$.
\end{itemize}
This finishes the construction of the instance $(D,\Delta)$ of \deltaExact\, which can clearly be done in polynomial time. We defer the correctness proof to the appendix.
\end{proof}

\end{comment}

%\subparagraph{Parameterized Hardness of \deltaExact.}\label{sec:w1hardness}

Next, we investigate the parameterized hardness of \deltaExact\ with respect to structural parameters of the underlying graph. We show that the problem is W[1]-hard when parameterized by the feedback vertex number of the underlying graph. 
The \emph{feedback vertex number} of a graph $G$ is the cardinality of a minimum vertex set $X\subseteq V(G)$ such that $G-X$ is a forest. The set $X$ is called a \emph{feedback vertex set}. 
Note that, in contrast to the previous results (\cref{thm:NPhardness}), the reduction we use to obtain the following result does not produce instances with a constant $\Delta$.

\begin{theorem}[$\star$]\label{thm:W1wrtFVS}
    \deltaExact\ is W[1]-hard when parameterized by the feedback vertex number of the underlying graph.
\end{theorem}
\begin{proof}
    We present a parameterized reduction from the W[1]-hard problem \textsc{Multicolored Clique} parameterized by the number of colors~\cite{fellows2009multipleinterval}.  Here, given a $k$-partite graph $H=(W_1\uplus W_2 \uplus\ldots\uplus W_k, F)$, we are asked whether $H$ contains a clique of size $k$. If $w\in W_i$, then we say that $w$ has \emph{color} $i$. W.l.o.g.\ we assume that $|W_1|=|W_2|=\ldots=|W_k|=n$ and that every vertex has at least one neighbor of every color. 
    Furthermore, for all $i\in[k]$, we assume the vertices in $W_i$ are ordered in some arbitrary but fixed way, that is, $W_i=\{w^i_1,w^i_2,\ldots,w^i_n\}$.
    Let $F_{i,j}$ with $i<j$ denote the set of all edges between vertices from $W_i$ and $W_j$. We assume w.l.o.g.\ that $|F_{i,j}|=m$ for all $i< j$.
    Furthermore, for all $i<j$ we assume that the edges in $F_{i,j}$ are ordered in some arbitrary but fixed way, that is, $F_{i,j}=\{e^{i,j}_1,e^{i,j}_2,\ldots,e^{i,j}_m\}$.

We give a reduction to a variant of \deltaExact\ where the period $\Delta$ is infinite (that is, there are no periods) and we allow $D$ to have infinity entries, meaning that the two respective vertices are not temporally connected. At the end of the proof we show how to obtain the result for a finite $\Delta$ and a matrix $D$ of durations of fastest paths, that only has finite entries.

    Given an instance $H$ of \textsc{Multicolored Clique}, we construct an instance $D$ of \deltaExact\ (with infinity entries and no periods) as follows. To ease the presentation, we first describe the underlying graph $G$ that is implicitly defined by the entries $D_{v,v'}=1$, that is, the pairs of vertices $v,v'$ that should be connected by a temporal path of duration one, meaning that there needs to be an edge connecting the two vertices. Afterwards, we describe the remaining entries of $D$.
    We will construct $G$ using several gadgets.

>>>>

\usepackage{caption} %for subfigure - join multiple figures and add captions
\usepackage{subcaption}
\usepackage{changepage} %inside figure we can put \begin{adjustwidth}{-1cm}{-1cm} \end{adjustwidth} and we can ignore the margins
\usepackage{mathtools} % write text [under]{over} arrow $\xrightarrow[\text{world}]{\text{hello}}$

@@

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Realizing temporal graphs from fastest travel times} %TODO Please add

@@

\begin{document}
\section{Ideas}
\begin{itemize}
    \item Symmetric distance matrix $D$: polytime?
    \item Use $|D|_\infty$ (maximum value in $D$) + max degree $\Delta$ as parameter $\rightarrow$ should give FPT
    \item FPT with feedback edge number
    \item use as parameter $\max_{i,j} |D_{i,j}-D_{j,i}|$
    \item approximation (additive or multiplicative), maybe with graph as input
\end{itemize}

\problemdef{\kDeltaExactLong\ (\kDeltaExact)}
{An $n \times n$ matrix $D$ of positive integers.}
{Does there exist a graph $G=(V,E)$ with vertices $\{v_1,\ldots,v_{n}\}$ 
and a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}^k$ such that, 
for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is equal to $D_{i,j}$.}

\problemdef{\kDeltaUpperBoundLong}
{An $n \times n$ matrix $D$ of positive integers and an integer $k\in \mathbb{N}$.}
{Does there exist a graph $G=(V,E)$ with vertices $\{v_1,\ldots,v_{n}\}$ with $n+k-1$ edges 
and a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ such that, 
for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is \emph{at most} $D_{i,j}$.}

\problemdef{\kDeltaLowerBoundLong}
{An $n \times n$ matrix $D$ of positive integers and an integer $k\in \mathbb{N}$.}
{Does there exist a graph $G=(V,E)$ on the vertices $\{v_1,\ldots,v_{n}\}$ with $n+k-1$ edges 
and a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ such that, 
for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is \emph{at least} $D_{i,j}$.}
\todo[inline]{NOT TEMPORALLY INTERESTING: the difficulty is to find a static graph with equal labels}

\todo[inline]{The matrix $D$ is m x m, and capturing the distances from an edge e to an edge f (zoom 14 Nov 22, GH).}
\todo[inline]{Check if the reduction works for the UPPER BOUNDED version (seems to work)}
\todo[inline]{how to extend the poly algorithms of cycles? Easy candidate: FES, FVS, series-parallel graphs, or tw=2 graphs}
\todo[inline]{FVS, $c$-factor approximation, hardness of approximation for $c<1.5$}
\end{document}
